\section*{Preface}

This book is a collection of notes for the ``Real and Functional Analysis'' (RFA, but universally known as ARF) course held in Politecnico di Milano, in the Mathematical Engineering Master's Degree.

It was written over the course of many years, with contributions by several Fubini-Tonelli authors: Gabriele Gabrielli, Teo Bucci, Gioele Cerri, Bruno Guindani.
We would also like to thank some external collaborators who helped in correcting some mistakes: Filippo Cipriani, Gabriele Corbo, Marco Lucchini.

This is also the first collection of lecture notes to be entirely written in English, which allows Fubini-Tonelli to expand in yet another area of knowledge, and hopefully among many new readers.
Unfortunately, the language barrier means that we will not be able to write as many bad jokes as we did in the Italian books, but we hope that readers will enjoy the journey nonetheless.

Real and Functional Analysis is a mandatory stepping stone for all future M.Sc. mathematical engineers, and it is notoriously one of the most difficult exams of this degree, perhaps \textit{the} most difficult one.
Therefore, we immediately warn the readers: it will \textit{not} be easy, and you \textit{will} spend sleep-deprived weekend nights trying to understand how the heck bidual Banach spaces interact with Ascoli-Arzelà's theorem.
But also, you \textit{will} encounter Fubini-Tonelli's theorem eventually, so there is really nothing to be worried about in the end.

\begin{flushright}
	\textit{The authors}\hspace*{0.5cm}
\end{flushright}